### 2023-03-17 chapters/paper_benchmarking.tex
Add review paper (WIP)
Change citation keys (WIP)

To advance the state of LVMs for speech, this paper (i) develops a benchmark for LVMs based on model likelihood, (ii) introduces a hierarchical LVM architecture without autoregressive decoder, (iii) compares LVMs to deterministic counterparts including WaveNet, and (iv) qualitatively and quantitatively evaluates the latent variables learned by different LVMs based on their usefulness for phoneme recognition. We find that:
\begin{itemize}[leftmargin=8mm]
    \vspace{-3mm}
    \setlength\itemsep{-0em}
    \item [(I)] State-of-the-art LVMs achieve likelihoods that are inferior to WaveNet at high temporal resolution but are superior at lower resolutions. Interestingly, we find that a standard LSTM \cite{hochreiter_long_1997} almost matches the likelihood of WaveNet.
    \item [(II)] LVMs with powerful autoregressive decoders achieve better likelihoods than the non-autoregressive LVM.
    \item[(III)] The expressiveness of LVMs for speech increases with a deeper hierarchy of stochastic latent variables, similar to conclusions within image modeling. % \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
    \item[(IV)] LVMs learn rich representations that are as good or better than Mel spectrograms for phoneme recognition also when using only 10 minutes of labeled data.
\end{itemize}\vspace{-3mm}
At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models
\footnote{\href{https://github.com/JakobHavtorn/benchmarking-lvms}{\url{github.com/JakobHavtorn/benchmarking-lvms}}}.
% \footnote{\fontsize{8.5}{0}\selectfont\texttt{ \url{github.com/anonymous/available-later}}}.
Before presenting the results, we provide a brief survey of existing LVMs for speech in a coherent notation.

@@

\subsection{Sequential deep latent variable models}
The selected models are all sequential deep latent variable models trained with variational inference and the reparameterization trick \cite{kingma_auto-encoding_2014}.
% The VRNN, SRNN, STCN and CW-VAE are all autoencoders trained with variational inference and the reparameterization trick \cite{kingma_auto-encoding_2014}. 
They take as input a variable-length sequence $\xb_{1:T} = \left(\xb_1, \xb_2, \dots, \xb_{T}\right)$ with $\xb_t \in \mathbb{R}^{D_x}$. 
% The VRNN, SRNN, STCN and CW-VAE are all VAEs \cite{kingma_auto-encoding_2014} that take as input a variable-length sequence $\xb_{1:T} = \left(\xb_1, \xb_2, \dots, \xb_{T}\right)$ with $\xb_t \in \mathbb{R}^{D_x}$. 
We let $\xb_{1:T}$ refer to the observed variable or a downsampled version of it. We will sometimes use $\xb$ to refer to the sequence $\xb_{1:T}$ when it is not ambiguous.
% In general, $\xb_{1:T}$ may refer to the original observed variable or a deterministically downsampled representation of the observed variable. We will sometimes use $\xb$ to refer to the full sequence $\xb_{1:T}$ when it is not ambiguous.

@@

            % \node[above=of s_2_1, yshift=-1.cm, xshift=2cm] (phi) {$p\pa{\xb_{1:T},\zb_{1:T}^{(1:L)}}$};
            \node[above=of s_2_1, yshift=-0.5cm, xshift=2.5cm] (phi) {CW-VAE generative};
        \end{tikzpicture}
    % \hfill
    }
\caption{
Inference (left) and generative (right) models for the CW-VAE with a hierarchy of $L=2$ latent variables, $s_1=1$ and $s_2=2$. 
The models are unrolled over four consecutive time steps but note that the graph continues towards $t=0$ and $t=T$. 
Blue arrows indicate parameter sharing between inference and generative models. 
We omit the deterministic variable of \cref{fig: autoregressive vrnn srnn cwvae 1L graphs} for clarity.
}
\label{fig: clockwork vae inference generative models 2 layered}
\vspace{-0mm}
\end{figure*}

@@

\section{Speech modeling benchmark}\label{sec:benchmark}
\paragraph{Data}
We train models on TIMIT \cite{garofolo_timit_1993}, LibriSpeech \cite{panayotov_librispeech_2015} and LibriLight \cite{kahn_libri-light_2020}. For TIMIT, we randomly sample 5\% of the training split for validation. 
We represent the audio as $\mu$-law encoded PCM standardized to values in $\bra{-1, 1}$ with discretization gap of ${2^{-b+1}}$. We use the original bit depth of $\SI{16}{bits}$ and sample rate of $\SI{16000}{Hz}$. We use this representation both as the input and the reconstruction target. We provide more details on the datasets in appendix~\cref{app: dataset details} and additional results on linear PCM in appendix~\cref{app: additional likelihood results appendix}.
%
\paragraph{Likelihood}
We report likelihoods in units of bits per frame (bpf) as this yields a more interpretable and comparable likelihood than total likelihood in nats. It also has direct connections with information theory and compression \cite{shannon_mathematical_1948, townsend_practical_2019}. In units of bits per frame, lower is better. For LVMs, we report the one-sample ELBO. The likelihoods can be seen in \cref{tab: likelihoods timit,tab: likelihoods librispeech}. We describe how to convert likelihood to bpf in appendix~\cref{app: likelihood in bits per frame}.
%
\paragraph{Models}
Architecture and training details are sketched below, while the full details are in appendices~\cref{app: model architectures} and \cref{app: training details} along with additional results for some alternative model configurations in appendix~\cref{app: additional likelihood results appendix}.
We select model configurations that can be trained on GPUs with a maximum of 12GB of RAM and train all models until convergence on the validation set.
We use a DMoL with 10 components for the output distribution of all models and model all datasets at their full bit depth of $\SI{16}{bits}$. We train and evaluate models on stacked waveforms similar to previous work \cite{chung_recurrent_2015, fraccaro_sequential_2016, aksan_stcn_2019} with stack sizes of $s=1$, $s=64$ and $s=256$. Hence, every model input $\xb_t$ is composed of $\tilde{x}_{t:t+s}$ where $\tilde{x}$ are waveform frames. We provide additional results with a Gaussian output distribution in appendix~\cref{app: additional likelihood results appendix}.

>>>>

To advance the state of LVMs for speech, this paper (i) develops a benchmark for LVMs based on model likelihood, (ii) introduces a hierarchical LVM architecture without autoregressive decoder, (iii) compares LVMs to deterministic counterparts including WaveNet, and (iv) qualitatively and quantitatively evaluates the latent variables learned by different LVMs based on their usefulness for phoneme recognition. We find that:
\begin{itemize}[leftmargin=8mm]
    \setlength\itemsep{-0em}
    \item [(I)] State-of-the-art LVMs achieve likelihoods that are inferior to WaveNet at high temporal resolution but are superior at lower resolutions. Interestingly, we find that a standard LSTM \cite{hochreiter_long_1997} almost matches the likelihood of WaveNet.
    \item [(II)] LVMs with powerful autoregressive decoders achieve better likelihoods than the non-autoregressive LVM.
    \item[(III)] The expressiveness of LVMs for speech increases with a deeper hierarchy of stochastic latent variables, similar to conclusions within image modeling. % \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
    \item[(IV)] LVMs learn rich representations that are as good or better than Mel spectrograms for phoneme recognition also when using only 10 minutes of labeled data.
\end{itemize}
At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models
\footnote{\href{https://github.com/JakobHavtorn/benchmarking-lvms}{\url{github.com/JakobHavtorn/benchmarking-lvms}}}.
% \footnote{\fontsize{8.5}{0}\selectfont\texttt{ \url{github.com/anonymous/available-later}}}.
Before presenting the results, we provide a brief survey of existing LVMs for speech in a coherent notation.

@@

\subsection{Sequential deep latent variable models}
The selected models are all sequential deep latent variable models trained with variational inference and the reparameterization trick \cite{kingma_autoencoding_2014}.
% The VRNN, SRNN, STCN and CW-VAE are all autoencoders trained with variational inference and the reparameterization trick \cite{kingma_auto-encoding_2014}. 
They take as input a variable-length sequence $\xb_{1:T} = \left(\xb_1, \xb_2, \dots, \xb_{T}\right)$ with $\xb_t \in \mathbb{R}^{D_x}$. 
% The VRNN, SRNN, STCN and CW-VAE are all VAEs \cite{kingma_auto-encoding_2014} that take as input a variable-length sequence $\xb_{1:T} = \left(\xb_1, \xb_2, \dots, \xb_{T}\right)$ with $\xb_t \in \mathbb{R}^{D_x}$. 
We let $\xb_{1:T}$ refer to the observed variable or a downsampled version of it. We will sometimes use $\xb$ to refer to the sequence $\xb_{1:T}$ when it is not ambiguous.
% In general, $\xb_{1:T}$ may refer to the original observed variable or a deterministically downsampled representation of the observed variable. We will sometimes use $\xb$ to refer to the full sequence $\xb_{1:T}$ when it is not ambiguous.

@@

            % \node[above=of s_2_1, yshift=-1.cm, xshift=2cm] (phi) {$p\pa{\xb_{1:T},\zb_{1:T}^{(1:L)}}$};
            \node[above=of s_2_1, yshift=-0.5cm, xshift=2.5cm] (phi) {CW-VAE generative};
        \end{tikzpicture}
    % \hfill
    }
\caption{
Inference (left) and generative (right) models for the CW-VAE with a hierarchy of $L=2$ latent variables, $s_1=1$ and $s_2=2$. 
The models are unrolled over four consecutive time steps but note that the graph continues towards $t=0$ and $t=T$. 
Blue arrows indicate parameter sharing between inference and generative models. 
We omit the deterministic variable of \cref{fig: autoregressive vrnn srnn cwvae 1L graphs} for clarity.
}
\label{fig: clockwork vae inference generative models 2 layered}
\end{figure*}

@@

\section{Speech modeling benchmark}\label{sec:glvm_benchmark}
\paragraph{Data}
We train models on TIMIT \cite{garofolo_timit_1993}, LibriSpeech \cite{panayotov_librispeech_2015} and LibriLight \cite{kahn_libri-light_2020}. For TIMIT, we randomly sample 5\% of the training split for validation. 
We represent the audio as $\mu$-law encoded PCM standardized to values in $\bra{-1, 1}$ with discretization gap of ${2^{-b+1}}$. We use the original bit depth of $\SI{16}{bits}$ and sample rate of $\SI{16000}{Hz}$. We use this representation both as the input and the reconstruction target. We provide more details on the datasets in appendix~\cref{app: dataset details} and additional results on linear PCM in appendix~\cref{app: additional likelihood results appendix}.
%
\paragraph{Likelihood}
We report likelihoods in units of bits per frame (bpf) as this yields a more interpretable and comparable likelihood than total likelihood in nats. It also has direct connections with information theory and compression \cite{shannon_mathematical_1948, townsend_practical_2019}. In units of bits per frame, lower is better. For LVMs, we report the one-sample ELBO. The likelihoods can be seen in \cref{tab: likelihoods timit,tab: likelihoods librispeech}. We describe how to convert likelihood to bpf in appendix~\cref{app: likelihood in bits per frame}.
%
\paragraph{Models}
Architecture and training details are sketched below, while the full details are in appendices~\cref{app: model architectures} and \cref{app: training details} along with additional results for some alternative model configurations in appendix~\cref{app: additional likelihood results appendix}.
We select model configurations that can be trained on GPUs with a maximum of 12GB of RAM and train all models until convergence on the validation set.
We use a DMoL with 10 components for the output distribution of all models and model all datasets at their full bit depth of $\SI{16}{bits}$. We train and evaluate models on stacked waveforms similar to previous work \cite{chung_recurrent_2015, fraccaro_sequential_2016, aksan_stcn_2019} with stack sizes of $s=1$, $s=64$ and $s=256$. Hence, every model input $\xb_t$ is composed of $\tilde{x}_{t:t+s}$ where $\tilde{x}$ are waveform frames. We provide additional results with a Gaussian output distribution in appendix~\cref{app: additional likelihood results appendix}.
